---
title: "Multi-omic analysis of adjusted mito results"
author: "David Amar"
date: "6/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# External libraries
library(data.table)
library(org.Hs.eg.db)
library(clusterProfiler)
library(org.Rn.eg.db)
library(MotrpacBicQC)
library(plotrix)
library(ggplot2)
library(gprofiler2)
library(corrplot)
# Load MoTrPAC data:
library(MotrpacRatTraining6moData)

mawg_repo = "~/Desktop/repos/motrpac-mawg/"
source(paste0(mawg_repo,'pass1b-06/integrative/clustering/cluster_viz_fx.R'))
data_dir = "~/Desktop/MoTrPAC/data/pass1b_6m/"
scratch = data_dir
knitr::opts_knit$set(root.dir = data_dir)
setwd(data_dir)
output_dir = "~/Desktop/repos/motrpac-rat-training-mitochondria/results/"
```

# Load datasets

Load mito biomarker data (see `mito_biomarker_analysis.Rmd`)
```{r}
load("~/Desktop/repos/motrpac-rat-training-mitochondria/data/mt_biomarkers.RData")
```

Load existing mito annotation data (see `mito_biomarker_analysis.Rmd`):
```{r}
load("~/Desktop/repos/motrpac-rat-training-mitochondria/data/mito_gene_annotation.RData")
```

Load DEA results from prot and RNA-seq

```{r}
rna_seq_dea = get(load(paste0(scratch,"ihw_adjusted_mito_paper_dea_results.RData")))
dea_combined = get(load(paste0(scratch,"ihw_adjusted_mito_paper_dea_prot_results.RData")))
dea_combined[["TRNSCRPT"]] = rna_seq_dea

# save supp data tables for the paper
for(ome in names(dea_combined)){
  curr_timewise = dea_combined[[ome]][["timewise_dea_mt_adjusted"]]
  curr_training = dea_combined[[ome]][["training_dea_mt_adjusted"]]
  curr_timewise$rname = paste(curr_timewise$feature_ID,curr_timewise$tissue_abbr)
  curr_training$rname = paste(curr_training$feature_ID,curr_training$tissue_abbr)
  curr_timewise = merge(curr_timewise,
      curr_training[,c("rname","p_value_male","p_value_female","p_value","ihw_q")],
      by="rname")
  curr_timewise = curr_timewise[,!colnames(curr_timewise)=="rname"]
  colnames(curr_timewise)[colnames(curr_timewise)=="p_value.y"] = "training_p_value"
  colnames(curr_timewise)[colnames(curr_timewise)=="p_value_male"] = "training_p_value_male"
  colnames(curr_timewise)[colnames(curr_timewise)=="p_value_female"] = "training_p_value_female"
  colnames(curr_timewise)[colnames(curr_timewise)=="ihw_q"] = "selection_fdr"
  # fwrite(
  #   curr_timewise,file=paste0(output_dir,ome,"_timeweise_dea.csv"),
  #   sep=",",quote = F,row.names = F
  # )
  timewise_df = curr_timewise
  save(timewise_df,file=paste0(output_dir,ome,"_timeweise_dea.RData"))
}

```

# Analyze the effect of adjustment

## Compare tissues by their adjustment effect on mito genes

```{r}
# Create a feature to gene to mito mapping
mt_symbols = mt$gene_symbol
# Remove RRBS and ATAC features
FEATURE_TO_GENE = FEATURE_TO_GENE[!grepl("cluster",FEATURE_TO_GENE$feature_ID),]
FEATURE_TO_GENE = FEATURE_TO_GENE[!grepl("^chr\\d",FEATURE_TO_GENE$feature_ID,perl=T),]
gc()
FEATURE_TO_GENE$is_mito = FEATURE_TO_GENE$gene_symbol %in% mt_symbols
mito_features = FEATURE_TO_GENE$feature_ID[FEATURE_TO_GENE$is_mito]
  
# TODO: think of another enrichment analysis as this may be biased
# # Initial comparison: do mt genes tend to have better robustness to adjustment?
# # TODO: add a conditional test, this test is biased by the underlying
# # number of mito genes in both sets (robust and not)
# mt_enrichment_training_level = c()
# for (ome in names(dea_combined)){
#   for(tissue in unique(dea_combined[[ome]]$training_dea$tissue_abbr)){
#     unadjusted = dea_combined[[ome]]$training_dea[
#       dea_combined[[ome]]$training_dea$tissue_abbr == tissue,
#     ]
#     adjusted = dea_combined[[ome]]$training_dea_mt_adjusted[
#       dea_combined[[ome]]$training_dea_mt_adjusted$tissue_abbr == tissue,
#     ]
#     unadjusted$is_mito = unadjusted$feature_ID %in% mito_features
#     adjusted$is_mito = adjusted$feature_ID %in% mito_features
#     rownames(adjusted) = adjusted$feature_ID
#     rownames(unadjusted) = unadjusted$feature_ID
#     adjusted = adjusted[rownames(unadjusted),]
#     training_summary_df = data.frame(
#       before = unadjusted$ihw_q < 0.05,
#       after = adjusted$ihw_q < 0.05,
#       is_mito = adjusted$is_mito
#     )
#     training_summary_df$robust = training_summary_df$before & training_summary_df$after
#     training_summary_df$not_robust = 
#       training_summary_df$before & !training_summary_df$after
#     training_summary_df$new = !training_summary_df$before & training_summary_df$after
#     
#     enrichment_res = c()
#     for(cc in c("robust","not_robust","new")){
#       enrichment_res[cc] = 1
#       tb = table(training_summary_df$is_mito,
#                 training_summary_df[[cc]])
#       if(is.null(dim(tb)) || length(tb)<3){next}
#       enrichment_res[cc] = fisher.test(tb,alternative = "g")$p.value
#     }
#     enrichment_res = data.frame(t(enrichment_res))
#     enrichment_res$tissue = tissue
#     enrichment_res$ome = ome
#     mt_enrichment_training_level = rbind(mt_enrichment_training_level,enrichment_res)
#   }
# }
# print(mt_enrichment_training_level)
```

## Merge the data from tissues with mito biomarker effects

```{r}
tissues_for_adjustment_comparisons = unique(
  mt_biomarker_dea$tissue[mt_biomarker_dea$kw_p < 0.05]
)
tissues_for_adjustment_comparisons = setdiff(
  tissues_for_adjustment_comparisons,
  c("TESTES","OVARY","VENACV")
)
tissues_for_adjustment_comparisons = gsub("-","",tissues_for_adjustment_comparisons)

get_zmat<-function(timewise,timewise_col=NULL){
  if(is.null(timewise_col)){
    timewise_col = "zscore"
  }
  if(!timewise_col %in% colnames(timewise)){
    timewise_col = "tscore"
  }
  rs = unique(timewise$feature_ID)
  cs = c(
    paste("female",sort(unique(timewise$comparison_group)),sep=","),
    paste("male",sort(unique(timewise$comparison_group)),sep=",")
  )
  m = matrix(NA,nrow=length(rs),ncol=length(cs),dimnames = list(rs,cs))
  for(col in cs){
    arr = strsplit(col,split=",")[[1]]
    tmp = timewise[
      timewise$sex == arr[1] & timewise$comparison_group == arr[2],
    ]
    m[tmp$feature_ID,col] = tmp[[timewise_col]] 
  }
  return(m)
}

tissue_adj_stats = c()
zmat = c()
fcmat = c()
adj_selected_sets = c()
for(ome in names(dea_combined)){
  for(tissue in tissues_for_adjustment_comparisons){
    if(!tissue %in% 
       unique(dea_combined[[ome]]$training_dea$tissue_abbr)){next}
    tr_before = dea_combined[[ome]]$training_dea[
      dea_combined[[ome]]$training_dea$tissue_abbr == tissue,
    ]
    tr_before = tr_before[tr_before$ihw_q < 0.05,]
    tr_before = tr_before[tr_before$feature_ID %in% mito_features,]
    
    tr_after = dea_combined[[ome]]$training_dea_mt_adjusted[
      dea_combined[[ome]]$training_dea_mt_adjusted$tissue_abbr == tissue,
    ]
    tr_after = tr_after[tr_after$ihw_q < 0.05,]
    tr_after = tr_after[tr_after$feature_ID %in% mito_features,]
    
    curr_features = union(tr_after$feature_ID,tr_before$feature_ID)
    if(length(curr_features)==0){next}
    feature_df = data.frame(
      feature_ID = curr_features,
      tissue = tissue,
      ome = ome,
      in_unadjusted = curr_features %in% tr_before$feature_ID,
      in_adjusted = curr_features %in% tr_after$feature_ID,
      multiomic_name = paste(ome,tissue,curr_features,sep=";")
    )
    adj_selected_sets = rbind(adj_selected_sets,feature_df)
    
    stats_df = data.frame(
      ome=ome,
      tissue = tissue,
      n_before = nrow(tr_before),
      n_after = nrow(tr_after),
      overlap = length(intersect(tr_before$feature_ID,tr_after$feature_ID)),
      union = length(curr_features),
      new = length(setdiff(tr_after$feature_ID,tr_before$feature_ID)),
      lost = length(setdiff(tr_before$feature_ID,tr_after$feature_ID))
    )
    stats_df$jaccard = stats_df$overlap/stats_df$union
    tissue_adj_stats = rbind(tissue_adj_stats,stats_df)
    
    # extract z-scores before vs. after for both sexes
    timewise_before = dea_combined[[ome]]$timewise_dea[
      dea_combined[[ome]]$timewise_dea$tissue_abbr == tissue,
    ]
    timewise_before = timewise_before[
      timewise_before$feature_ID %in% curr_features,
    ]
    z_before = get_zmat(timewise_before)
    colnames(z_before) = paste(colnames(z_before),"before",sep=",")
    fc_before = get_zmat(timewise_before,"logFC")
    colnames(fc_before) = paste(colnames(fc_before),"before",sep=",")
    
    timewise_after = dea_combined[[ome]]$timewise_dea_mt_adjusted[
      dea_combined[[ome]]$timewise_dea_mt_adjusted$tissue_abbr == tissue,
    ]
    timewise_after = timewise_after[
      timewise_after$feature_ID %in% curr_features,
    ]
    z_after = get_zmat(timewise_after)
    colnames(z_after) = paste(colnames(z_after),"after",sep=",")
    fc_after = get_zmat(timewise_after,"logFC")
    colnames(fc_after) = paste(colnames(fc_after),"before",sep=",")
    
    if(nrow(z_before)>1){
      merged_zs = cbind(z_before,z_after[rownames(z_before),])
      merged_fc = cbind(fc_before,fc_after[rownames(fc_before),])
    }
    else{
      merged_zs = cbind(z_before,z_after)
      merged_fc = cbind(fc_before,fc_after)
    }
    
    rownames(merged_zs) = paste(ome,tissue,rownames(merged_zs),sep=";")
    zmat = rbind(zmat,merged_zs)
    rownames(merged_fc) = paste(ome,tissue,rownames(merged_fc),sep=";")
    fcmat = rbind(fcmat,merged_fc)
  }
}
rownames(adj_selected_sets) = adj_selected_sets$multiomic_name

tissue_adj_stats = tissue_adj_stats[order(tissue_adj_stats$jaccard),]
tissue_adj_stats$col = tissue_cols[tissue_adj_stats$tissue]
tissue_adj_stats$ord = 1:nrow(tissue_adj_stats)
tissue_adj_stats$name = paste(tissue_adj_stats$ome,tissue_adj_stats$tissue,sep=";")

ggplot(
    tissue_adj_stats, 
    aes(y=log10(1+n_before), 
    x=log10(1+n_after),label=tissue,col=jaccard,shape=ome)) + 
    geom_point(size=2) + 
    ylab("log10 selected, 5% FDR, unadjusted") + 
    xlab("log10 selected, 5% FDR, adjusted") + geom_abline(slope=1) + 
    xlim(0,0.2+max(log10(1+tissue_adj_stats$n_after))) +
    geom_text(hjust=-0.1, vjust=0,size=4)

```

## Analysis of the timewise z-scores

```{r}
z_diffs = zmat[,1:8] - zmat[,9:16]
par(mfrow = c(2,4))
for(j in 1:8){hist(abs(z_diffs[,j]),
  breaks=50,xlim=c(0,8),
  main=gsub(",before","",colnames(zmat)[j]))}
z_diff_features = apply(abs(z_diffs)>3,1,any)
table(z_diff_features)
table(z_diff_features) / length(z_diff_features)
table(adj_selected_sets[names(which(z_diff_features)),"in_adjusted"])
table(adj_selected_sets[names(which(z_diff_features)),"in_unadjusted"])

```

```{r}
k_res = list()
set.seed(123)
clustering_analysis_matrix = zmat
for(k in 1:20){
  k_res[[k]] = kmeans(clustering_analysis_matrix[z_diff_features,],
                      k,nstart = 50,iter.max = 500)
}
within_rss = sapply(k_res,function(x)sum(x$withinss))
plot(within_rss,type="b",ylab = "Within cluster SS",xlab="Number of clusters",col="blue")
arrs = strsplit(rownames(clustering_analysis_matrix)[z_diff_features],
                split=";")
arrs = t(sapply(arrs,function(x)x))

cluster_obj = k_res[[4]]$cluster
cluster_res = data.frame(
  cluster = cluster_obj,
  ome=arrs[,1],feature_ID = arrs[,3],tissue=arrs[,2]
)
cluster_res$feature_raw = names(cluster_obj)
cluster_res$feature = gsub("SKMGN","SKM-GN",cluster_res$feature_raw)
cluster_res$feature = gsub("SKMVL","SKM-VL",cluster_res$feature)
cluster_res$tissue =gsub("SKMGN","SKM-GN",cluster_res$tissue)
cluster_res$tissue = gsub("SKMVL","SKM-VL",cluster_res$tissue)

clustering_analysis_matrix1 = clustering_analysis_matrix[z_diff_features,1:8]
clustering_analysis_matrix1_cols = t(sapply(
  strsplit(colnames(clustering_analysis_matrix1),split=","),function(x)x))
colnames(clustering_analysis_matrix1) = 
  paste(clustering_analysis_matrix1_cols[,1],clustering_analysis_matrix1_cols[,2],sep="_")
pls_before = plot_cluster_trajectories(
  clustering_analysis_matrix1[,1:8],
  cluster_res[,setdiff(colnames(cluster_res),"feature_raw")])

clustering_analysis_matrix2 = clustering_analysis_matrix[z_diff_features,9:16]
clustering_analysis_matrix2_cols = 
  t(sapply(strsplit(colnames(clustering_analysis_matrix2),split=","),function(x)x))
colnames(clustering_analysis_matrix2) = 
  paste(clustering_analysis_matrix2_cols[,1],clustering_analysis_matrix2_cols[,2],sep="_")
pls_after = plot_cluster_trajectories(
  clustering_analysis_matrix2[,1:8],
  cluster_res[,setdiff(colnames(cluster_res),"feature_raw")])


library(gridExtra)
pls = list()
for(i in 1:length(pls_before[[2]])){
  pls = c(pls,list(pls_before[[2]][[i]],pls_after[[2]][[i]]))
}
grid.arrange(grobs = pls, ncol=4)

table(cluster_res$cluster,cluster_res$ome)
features_per_cluster(cluster_res)

clust_res_supp_df = cluster_res
rownames(clust_res_supp_df)  = clust_res_supp_df$feature_raw
curr_f2g = FEATURE_TO_GENE[
  FEATURE_TO_GENE$feature_ID %in% clust_res_supp_df$feature_ID,
]
clust_res_supp_df = cbind(clust_res_supp_df,
  clustering_analysis_matrix[rownames(clust_res_supp_df),])
clust_res_supp_df = merge(
  clust_res_supp_df,
  curr_f2g[,c("feature_ID","gene_symbol")],
  by.x = "feature_ID",
  by.y = "feature_ID"
)
colnames(clust_res_supp_df) = gsub(
  ",","_",
  colnames(clust_res_supp_df)
)
write.csv(
  clust_res_supp_df,
  file = paste0(output_dir,"supp_clustering_results.csv"),
  row.names = F,quote = F
)

# create a clustering solution by tissue for downstream analysis
selected_sets = list()
for(cl in unique(cluster_obj)){
  cl_features = names(cluster_obj)[cluster_obj==cl]
  arrs = strsplit(cl_features,split=";")
  cl_ome = sapply(arrs,function(x)x[1])
  cl_tissue = sapply(arrs,function(x)x[2])
  cl_f_ids = sapply(arrs,function(x)x[3])
  tissue_size = table(cl_tissue)
  curr_tissues = names(tissue_size)[tissue_size>9]
  for(tissue in curr_tissues){
    curr_name = paste(tissue,cl,sep="_")
    curr_features = cl_f_ids[cl_tissue==tissue]
    curr_genes = FEATURE_TO_GENE[
      FEATURE_TO_GENE$feature_ID %in% curr_features,"gene_symbol"]
    selected_sets[[curr_name]] = curr_genes
  }
}
sort(sapply(selected_sets,length))

```

## Run enrichment analysis

```{r}

simple_enrichment_analysis<-function(s,univ,pathways,min_size=5,setname=""){
  s = intersect(s,univ)
  if(length(s)<min_size){
    print("Warning: input set is smaller than min_size, returning NULL")
    return(NULL)
  }
  v1 = univ %in% s
  enrichment_results = c()
  for(pname in names(pathways)){
    pg = pathways[[pname]]
    pg = intersect(pg,univ)
    if(length(pg)<min_size){next}
    v2 = univ %in% pg
    fisher_res = fisher.test(table(v1,v2),alternative = "g")
    overlap_set = intersect(s,pg)
    res = data.frame(
        setname=setname,
        pathway = pname,
        setsize = length(s),pathwaysize=length(pg),
        overlap = length(overlap_set),
        p = fisher_res$p.value,
        genes = paste(overlap_set,collapse=",")
    )
    enrichment_results = rbind(enrichment_results,res)
  }
  return(enrichment_results)
}

enrichment_results = c()
for (setname in names(selected_sets)){
  tissue = strsplit(setname,split="_")[[1]][1]
  if(tissue=="WATSC"){tissue="WAT-SC"}
  if(tissue=="SKMGN"){tissue="SKM-GN"}
  if(tissue=="SKMVL"){tissue="SKM-VL"}
  univ = GENE_UNIVERSES$gene_symbol$TRNSCRPT[[tissue]]
  univ = univ[univ %in% mt$gene_symbol]
  enr_res =  simple_enrichment_analysis(
              selected_sets[[setname]],univ,
              mitocarta_pathways,setname = setname)
  if(!is.null(enr_res)){
    enrichment_results=rbind(enrichment_results,enr_res)
  }
}
enrichment_results$q = p.adjust(enrichment_results$p,method="fdr")
selected_enrichment_results = enrichment_results[enrichment_results$q < 0.1 &
  enrichment_results$overlap > 1 & enrichment_results$pathwaysize < 200,]
dim(selected_enrichment_results)

selected_enrichment_results[,c("setname","pathway","q","genes")]
write.table(
  selected_enrichment_results,
  file = paste0(output_dir,"selected_mitocarta_enrichment_results.tsv"),
  row.names = F,quote = F,sep="\t")

```

```{r}
# Standard enrichment (not mitocarta)
library(gprofiler2)
res = gost(selected_sets,
    organism='rnorvegicus',
    significant=FALSE, # return all results
    correction_method='fdr', # it will not return unadjusted p-values. 
    sources=c('KEGG','REAC'),
    custom_bg=mt$gene_symbol,
    domain_scope='custom',
    evcodes=T)
pathway_enrichments = res$result
pathway_enrichments$q = p.adjust(pathway_enrichments$p_value,method="fdr")
table(pathway_enrichments$q < 0.1)
selected_pathway_enrichment_results = pathway_enrichments[
  pathway_enrichments$q < 0.1 &
  pathway_enrichments$intersection_size > 1 & pathway_enrichments$term_size < 200,]
dim(selected_pathway_enrichment_results)
selected_pathway_enrichment_results$term_name[
  grepl(
    "Respiratory electron transport, ATP synthesis by chemiosmotic coupling,",
    selected_pathway_enrichment_results$term_name
  )
] = "Electron transport,ATP synthesis,heat production"

selected_pathway_enrichment_results[,c("query","term_name","q","intersection")]
pathway_enrichment_columns = c(
  "query","term_name","p_value","q",
  "term_size","query_size","intersection_size","precision","recall",
  "effective_domain_size","source","term_id","intersection"
)
write.table(
  selected_pathway_enrichment_results[,pathway_enrichment_columns],
  file = paste0(output_dir,"selected_pathway_enrichment_results.tsv"),
  row.names = F,quote = F,sep="\t")
```



